Computing Vowel Elision in Yoruba Collocations

### 2023-04-20 paper.tex

The phenomenon of vowel hiatus in Yoruba, a language in the Niger-Congo family
spoken primarily in West Africa, is well-documented. A large body of work
exists that explores the various aspects of it, largely from an
optimality-theoretic perspective \cite{ola2002yoruba, seidl2000yoruba}. Two
particularly well-studied instances of this problem are \textit{assimilation}
and \textit{deletion}. These are the two primary strategies employed by the
language for resolving a sequence of two or more vowels that occur at the
boundary of two adjoined morphemes. This text will specifically explore the
second strategy of deletion, and in doing so, will attempt to provide an
analysis that is simple enough to be feasible to verify computationally. That
is, once we have developed our analysis, the core contribution of this work
will be to encode it computationally so it can be quickly and easily verified
whether it works for a range of corpus data. We employ an optimality-theoretic
perspective throughout, and provide a thorough discussion in Section
\ref{sec:computing-optimal-surface-forms} of the prior work on computational
interpretations of optimality theory. Afterwards, we describe our own approach
to automating our analysis and discuss its efficacy.

@@

We can describe these three distinct groups of data straightforwardly. In
Yoruba, when two vowels become adjacent via concatenation of Yoruba morphemes,
the first vowel is deleted, unless the second vowel is less sonorous,
in which case it is deleted. This deletion occurs in order to ensure that
syllables always have an onset. This generalization can be analyzed from an
Optimality-theoretic viewpoint. To begin, the phenomena in
\pref{fig:deletion-simple} can be naturally accounted for with a prohibition
against onsetless syllables, so one of the conjoined vowels is deleted to
prevent such a syllable from surfacing. Specifically, a `V`, with any number of
`C`s after it, is not a valid candidate for a syllable, because it lacks an
onset. We can codify this observation via the constraint \onset, which
precisely moderates against candidate surface forms with syllables which lack
an onset. This is defined in \pref{def:onset-and-maxc}.

% \textbf{make sure we're formulating the foot binarity effect correctly}

% \textbf{cite Yoruba Vowel Elisision: Minimality Effects}

@@

This helps inform why, for the data in \pref{fig:deletion-except-1}, [i] is
always elided, even though it appears morpheme-initially underlyingly.
Specifically, the language prefers to delete less sonorous vowels such as
[i] over more sonorous vowels. We can codify this intuition into a faithfulness
constraint, \ssp. This constraint is defined in \pref{def:ssp}.

@@

We notice that candidate three is a winner threatening loser! It deletes both
of the non-sonorous vowels at the boundary of the two morphemes, and instead
epenthesizes an [a], which is a very sonorous vowel, incurring no violations of
\ssp. In order to prevent this winner-threatening loser, we introduce a
high-ranked \dep{} constraint. This is supported by our data, because we never
see epenthesis. We now show a simple ranking tableau that represents this argument.

@@

\label{sec:computing-optimal-surface-forms}
\section{Computing Optimal Surface Forms}

@@

Our first order of business is to generate a reasonably large set of
permutations of \textipa{[sol\'u]}, one of which is the correct output form. We
are not yet attempting to identify which is the correct output form, only
generate it so it may be picked later. In effect, we are attempting to
efficiently emulate Optimality Theory's \textsc{Gen}. To do this, we intoduce
the following core transformations; we model each of these transformations as
functions of their own:

@@

\begin{figure}[h]
\caption{Our representation of words in Yoruba, along with our strategy for deletion.}
\label{fig:impl-deletion}
\begin{minted}{rust}
#[derive(Debug, Clone)]
struct SyllabifiedCandidate {
    form: Vec<Segment>,
    rng: StdRng,
}

@@

We now describe key details of our implementation. The full source code may be
found at \href{https://github.com/hemantgouni/linguistics-capstone}{github.com/hemantgouni/linguistics-capstone}.
Starting with the basics, Figure \ref{fig:impl-deletion} shows both our
representation of Yoruba words, and our simple handling of deletion.
Specifically, a word is represented as a list of segments (this is the meaning of
\texttt{Vec<Segment>}). At this point, you might ask: why not represent words simply
as a list of characters? What's the difference between a character and a segment?
This reflects a complication in computers' handling of text beyond simple English
characters, covered by the ASCII character encoding. Unicode, an improvement of
ASCII that handles many more languages beyond those which use the Latin alphabet,
would consider the accent mark and letter \textipa{O} in \textipa{\'O}, to be
separate characters, and not part of the same. We must sidestep this issue by
parsing words into \textit{graphemes}, which accounts for the intuitive notion
of what a character is.

@@

        onsets - syllabi
    }
}
\end{minted}
\end{figure}

The other significant portion of our implementation, beyond generating
candidates, is evaluating them. This, of course, is done with constraints.
Figure \ref{fig:impl-onset} shows our implementation of the \onset{}
markedness constraint. On lines 2-7, we count the number of syllable
nuclei in the input (a \texttt{SyllabifiedCandidate} from Figure
\ref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
Onsets and nuclei are calculated, along with other linguistic data, upon
the creation of a \texttt{SyllabifiedCandidate}; we merely analyze them
here. The number of violations of \onset{} is given by the number of
onsets minus the number of nuclei. This will never be a negative number,
because the algorithm that marks onsets and nuclei will only create one
onset per nucleus, at most (because this is the maximum allowed by
Yoruba).

@@

@@

TODO: Requires implementation to be reasonably far along first!

>>>>

The phenomenon of vowel hiatus in Yoruba, a language in the Niger-Congo family
spoken primarily in West Africa, is well-documented. A mature body of work
exists that explores the various aspects of it, largely from an
optimality-theoretic perspective \cite{ola2002yoruba, seidl2000yoruba,
pulleyblank1988vowel}. Two particularly well-studied instances of this problem
are the \textit{assimilation} case and the \textit{deletion} case. These are the
two primary strategies employed by the language for resolving sequences of two
or more vowels that occur at the boundary of two adjoined morphemes
(sequences of two vowels do not generally appear in Yoruba words due to
syllable structure constraints). This text will specifically explore the
second strategy of deletion, and in doing so, will specially attempt to provide
an analysis that is simple enough to be feasible to verify computationally. That
is, once we have developed our analysis, the core contribution of this work
will be to encode it as a computer program so it can be quickly and easily
verified whether it works for a range of corpus data. We employ an
optimality-theoretic perspective throughout, and provide a thorough discussion
in Section \ref{sec:computing-optimal-surface-forms} of the prior work on
computational interpretations of optimality theory. Afterwards, we describe
our own approach to automating our analysis and discuss its efficacy.

@@

We can describe these three distinct groups of data straightforwardly. In
Yoruba, when two vowels become adjacent via concatenation of Yoruba morphemes,
the first vowel is deleted, unless the second vowel is less sonorous,
in which case it is deleted (the Yoruba vowel inventory and sonority scale is
discussed on page 4). This deletion occurs in order to ensure that syllables
always have an onset. This generalization can be analyzed from an
Optimality-theoretic viewpoint. To begin, the phenomena in
\pref{fig:deletion-simple} can be naturally accounted for with a prohibition
against onsetless syllables, so one of the conjoined vowels is deleted to
prevent such a syllable from surfacing. Specifically, a `V`, with any number of
`C`s after it, is not a valid candidate for a syllable, because it lacks an
onset. We can codify this observation via the constraint \onset, which
precisely moderates against candidate surface forms with syllables which lack
an onset. This is defined in \pref{def:onset-and-maxc}.

@@

This helps inform why, for the data in \pref{fig:deletion-except-1}, [i] is
always elided, even though it appears morpheme-initially underlyingly.
Specifically, the language prefers to delete less sonorous vowels such as
[i] over more sonorous vowels. We can codify this intuition into a faithfulness
constraint, \ssp{} (which stands for `Sonority Sequencing Principle').
This constraint is defined in \pref{def:ssp}.

@@

We notice that candidate four is a winner threatening loser! It deletes both
of the non-sonorous vowels at the boundary of the two morphemes, and instead
epenthesizes an [a], which is a very sonorous vowel, incurring no violations of
\ssp. In order to prevent this winner-threatening loser, we introduce a
high-ranked \dep{} constraint. This is supported by our data, because we never
see epenthesis. We now show a simple ranking tableau that represents this argument.

@@

\section{Computing Optimal Surface Forms}
\label{sec:computing-optimal-surface-forms}

@@

Our first order of business is to generate a reasonably large set of
permutations of \textipa{/se-ol\'u/}, one of which is the correct output form.
We are not yet attempting to identify which is the correct output form, only
generate it so it may be picked later. In effect, we are attempting to
efficiently emulate Optimality Theory's \textsc{Gen}. To do this, we intoduce
the following core transformations; we model each of these transformations as
functions of their own:

@@

\begin{figure}[h]
\caption{Our representation of words in Yoruba, along with our strategy for deletion.}
\label{fig:impl-deletion}
\begin{minted}[linenos]{rust}
struct SyllabifiedCandidate {
    form: Vec<Segment>,
    rng: StdRng,
}

@@

We now describe key details of our implementation. The full source code may be
found at \href{https://github.com/hemantgouni/linguistics-capstone}{github.com/hemantgouni/linguistics-capstone}.
Starting with the basics, Figure \ref{fig:impl-deletion} shows both our
representation of Yoruba words and our simple handling of deletion.
Specifically, a word is represented as a list of segments; this is denoted by
\texttt{Vec<Segment>} on line 2.
At this point, it is reasonable to ask: why not represent words simply
as a list of characters? What's the difference between a character and a segment?
This reflects a complication in computers' handling of text beyond the
simple English characters covered by the ASCII character encoding. Unicode, an
improvement of ASCII that handles many more languages beyond those which use the
Latin alphabet, would consider the accent mark and letter \textipa{O} in \textipa{\'O},
to be separate characters, and not part of the same. We must sidestep this issue by
parsing words into \textit{graphemes}, which accounts for the intuitive notion
of what a character is: \textipa{\'O} is a single grapheme.

@@

        syllabi - onsets
    }
}
\end{minted}
\end{figure}

The other significant portion of our implementation, beyond generating
candidates, is evaluating them. This, of course, is done with constraints.
Figure \ref{fig:impl-onset} shows our implementation of the \onset{}
markedness constraint. On lines 2-7, we count the number of syllable
nuclei in the input (a \texttt{SyllabifiedCandidate} from Figure
\ref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
Onsets and nuclei are calculated, along with other linguistic data, upon
the creation of a \texttt{SyllabifiedCandidate}; we merely analyze them
here. The number of violations of \onset{} is given by the number of
nuclei minus the number of onsets. This will never be a negative number,
because the algorithm that marks onsets and nuclei will only create one
onset per nucleus, at most (because this is the maximum allowed by
Yoruba).

@@

For the sake of simplicity, we'll perform a single deletion operation.
To do this, we must parse \textipa{/ow\'o-k\'i-ow\'o/} from a string
into a \texttt{SyllabifiedCandidate}. We split it into each of its
graphemes, creating the list \{\textipa{o,w,\'o,k,\'i,o,w,\'o}\}. To
perform a deletion, we simply choose a random element of this list,
say, [k]. This yields \{\textipa{o,w,\'o,\'i,o,w,\'o}\}.

To find the number of \onset{} violations, we note that our list is
annotated, internally, as \{\texttt{Nucleus}, \texttt{Onset},
\texttt{Nucleus}, \texttt{Nucleus}, \texttt{Nucleus}, \texttt{Onset},
\texttt{Nucleus}\}. Counting the number of nuclei, we find that there
are 5. There are only two onsets. 5 - 2 = 3, so there are 3
violations of \onset.

Finally, to count the number of \dep{} violations, we turn
\{\textipa{o,w,\'o,\'i,o,w,\'o}\} back into the string
[\textipa{ow\'o\'iow\'o}]. We then use a diff algorithm to determine
the sequence of edit operations that could be used to arrive at this
string from \textipa{/ow\'o-k\'i-ow\'o/}. In this case, it tells us
that a single \texttt{Delete} is all that is necessary, as expected.
This does not cause any violations of \dep{} because it is not an
\texttt{Insert} operation, so we return 0 violations.

@@

TODO: Do the evaluations now that the implementation is reasonably
complete!
